---
title: "dotplots"
output: github_document
---

## Libraries and setup

```{r setup, warning = FALSE, message = FALSE}
library(tidyverse)
library(cowplot)

theme_set(theme_cowplot())
```


## Read in data

Here are predictions from a simple model from [https://covidactnow.org/](https://covidactnow.org/) as of March 24:

```{r data}
df = read_csv("covid_act_now.csv", col_types = cols(
  date = col_date(),
  not_flattened = col_integer(),
  flattened = col_integer(),
  beds = col_integer()
))

df
```

And a long-format version for hospitalizations:

```{r hosp}
hosp = df %>%
  pivot_longer(c(not_flattened, flattened), names_to = "model", values_to = "hospitalized") %>%
  mutate(flattened = model == "flattened")

hosp
```


## Area chart version

Area chart version folks tend to use:

```{r area_chart}
hosp %>%
  ggplot(aes(x = date)) +
  geom_area(
    aes(y = hospitalized, color = flattened, fill = flattened),
    position = "identity", alpha = 0.5
  ) +
  geom_line(aes(y = beds), data = df)
```


## Dotplots

Let's try to re-generate this using a frequency framing approach with dotplots. First we'll pick a dot size in terms of number of people:

```{r people_per_dot}
people_per_dot = 10000
```

Then we'll round counts accordingly:

```{r make_dots}
hosp_dots = hosp %>%
  mutate(
    # TODO: need a better solution for the very small numbers
    hospitalized_dots = ceiling(hospitalized / people_per_dot),
    dot_number = map(hospitalized_dots, ~ seq_len(.x))
  ) %>%
  unnest(dot_number) %>%
  mutate(
    has_a_bed = dot_number * people_per_dot < beds
  )

hosp_dots
```

Then we'll draw as dots:

```{r dotplot_combined, fig.width = 6, fig.height = 4}
bin_width = 2
dodge_width = 3.5

hosp_dots %>%
  ggplot(aes(x = date)) +
  geom_area(aes(y = beds), data = df, fill = "gray85") +
  geom_dotplot(
    aes(fill = flattened), color = NA, binwidth = bin_width, alpha = 0.5, 
    position = position_dodge(width = dodge_width)
  ) +
  geom_dotplot(
    aes(group = flattened), fill = "gray50", color = NA,
    binwidth = bin_width,
    position = position_dodge(width = dodge_width),
    data = . %>% filter(has_a_bed)
  ) +
  annotate("text",
    x = as.Date("2020-05-25"), y = max(df$beds) + 10000, 
    label = "Has a bed", fontface = "bold", hjust = 0, vjust = 0,
    color = "gray65"
  ) +
  annotate("text",
    x = as.Date("2020-04-25"), y = max(df$flattened), 
    label = "Doesn't have a bed", fontface = "bold", hjust = 0, vjust = 0,
    color = "black"
  ) +
  scale_fill_brewer(palette = "Set1") +
  # TODO: scaling of the area chart is slightly wrong here
  # because of how dotplots work in ggplot, probably could
  # manually lay out dots instead (wouldn't be hard using
  # hosp_dots$dot_number)
  scale_y_continuous(limits = c(0, max(hosp$hospitalized)))
```

Or the facetted version:

```{r dotplot_faceted, fig.width = 6, fig.height = 8}
hosp_dots %>%
  ggplot(aes(x = date)) +
  geom_area(aes(y = beds), data = hosp, fill = "gray85") +
  geom_dotplot(
    aes(fill = flattened), color = NA, binwidth = bin_width, alpha = 0.5, 
    position = position_dodge(width = dodge_width)
  ) +
  geom_dotplot(
    aes(group = flattened), fill = "gray50", color = NA,
    binwidth = bin_width,
    position = position_dodge(width = dodge_width),
    data = . %>% filter(has_a_bed)
  ) +
  annotate("text",
    x = as.Date("2020-05-25"), y = max(df$beds) + 10000, 
    label = "Has a bed", fontface = "bold", hjust = 0, vjust = 0,
    color = "gray65"
  ) +
  annotate("text",
    x = as.Date("2020-04-25"), y = max(df$flattened), 
    label = "Doesn't have a bed", fontface = "bold", hjust = 0, vjust = 0,
    color = "black"
  ) +
  scale_fill_brewer(palette = "Set1", guide = FALSE) +
  scale_y_continuous(limits = c(0, max(hosp$hospitalized))) +
  facet_grid(flattened ~ .)
```

